---
title: "Tree Viewer"
description: |
  Interactive visualization of FTOL
title-block-banner: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

tree_url <- "https://taxonium.org/?treeUrl=https%3A%2F%2Fraw.githubusercontent.com%2Ffernphy%2Fftol_data%2Fmain%2Fftol_sanger_con_dated.tre&ladderizeTree=true&treeType=nwk&metaUrl=https%3A%2F%2Fraw.githubusercontent.com%2Ffernphy%2Fftol_data%2Fmain%2Fftol_sanger_sampling.csv&xType=x_dist&color=%7B%22field%22%3A%22meta_family%22%7D&srch=%5B%5D" # nolint
```

The interactive tree visualization below shows higher taxonomic groups (genus, family, order, etc.) by color on the dated, consensus tree (FTOL v`r ftolr::ft_data_ver()`). Default coloring is by family. Species names will appear when zoomed in or hovering over a tip. The same visualization is available at [Taxonium](`r tree_url`), where you can also upload other trees for viewing. For more details about using Taxonium, see the [documentation](https://docs.taxonium.org/en/latest/index.html#).

If you want to view the tree in a desktop app, we recommend either [Dendroscope3](https://uni-tuebingen.de/fakultaeten/mathematisch-naturwissenschaftliche-fakultaet/fachbereiche/informatik/lehrstuehle/algorithms-in-bioinformatics/software/dendroscope/) or the [Taxonium desktop app](https://docs.taxonium.org/en/latest/app.html).

```{r, out.width=720, out.extra='style="border: 1px solid #464646;" allowfullscreen="" allow="autoplay"'}
knitr::include_url(
  tree_url,
  height = 1200)
```